---
title: "ğ‘·ğ’“ğ’†ğ’…ğ’Šğ’„ğ’•ğ’Šğ’ğ’ˆ ğ‘ªğ’‰ğ’“ğ’ğ’ğ’Šğ’„ ğ‘«ğ’Šğ’”ğ’†ğ’‚ğ’”ğ’† ğ‘¹ğ’Šğ’”ğ’Œ ğ’Šğ’ ğ‘ªğ’‰ğ’Šğ’ğ’…ğ’“ğ’†ğ’ ğ‘¼ğ’”ğ’Šğ’ğ’ˆ ğ‘´ğ’‚ğ’„ğ’‰ğ’Šğ’ğ’† ğ‘³ğ’†ğ’‚ğ’“ğ’ğ’Šğ’ğ’ˆ"
author: "Niteesha Adapala"
---

# Filtering and Summarizing Relevant Variables from a Dataset

```{r setup,echo=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(readr)
library(readxl)

data_path <- "C:/Users/Niteesha/Desktop/big data/NSCH2016_Topical_SPSS_CAHM_DRCv2.csv"
variables_path <- "C:/Users/Niteesha/Desktop/big data/2016NSCH_Variable name_labels.xlsx"

data <- read_csv(data_path)
variables <- read_excel(variables_path, sheet = 1)

colnames(variables) <- c("Variable_Name", "Description")

keywords <- c("disease", "health", "mental", "development", "disability", 
              "chronic", "condition", "anxiety", "depression", "ADHD", "autism")

relevant_vars <- variables %>%
  filter(grepl(paste(keywords, collapse = "|"), Description, ignore.case = TRUE))

cat("Relevant Variables:\n")
print(relevant_vars)

selected_data <- data %>%
  select(all_of(relevant_vars$Variable_Name))

cat("\nSummary of Selected Data:\n")
print(summary(selected_data))

cat("\nMissing Values:\n")
print(colSums(is.na(selected_data)))

```

# Variable Filtering and Feature Selection for Dataset Preparation

```{r,echo=FALSE, warning=FALSE, message=FALSE}
individual_factors <- c("age", "gender", "race", "ethnicity", "BMI", 
                        "physical_activity", "time_on_devices", "screen_time")
clinical_conditions <- c("chronic_condition", "comorbidities", "mental_health", 
                         "depression", "anxiety", "autism", "ADHD")
family_environment <- c("family_income", "smoking_environment", 
                        "neighborhood_safety", "social_support")

selected_individual_vars <- variables %>%
  filter(grepl(paste(individual_factors, collapse = "|"), Description, ignore.case = TRUE))

selected_clinical_vars <- variables %>%
  filter(grepl(paste(clinical_conditions, collapse = "|"), Description, ignore.case = TRUE))

selected_family_vars <- variables %>%
  filter(grepl(paste(family_environment, collapse = "|"), Description, ignore.case = TRUE))

selected_vars <- bind_rows(selected_individual_vars, selected_clinical_vars, selected_family_vars)

final_selected_vars <- selected_vars %>%
  distinct(Variable_Name, .keep_all = TRUE) %>%
  slice_head(n = 20)  # Adjust number if needed
final_data <- data %>%
  select(all_of(final_selected_vars$Variable_Name))

cat("Final Selected Variables:\n")
print(final_selected_vars)

cat("\nSummary of Final Data:\n")
print(summary(final_data))
```
# Data Cleaning: Handling Missing Values and Data Transformation

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(tidyverse) 

final_data_cleaned <- final_data %>%
  mutate(across(where(is.numeric), ~ ifelse(is.na(.), mean(., na.rm = TRUE), .))) %>%
  drop_na()

str(final_data_cleaned)
final_data_cleaned <- final_data_cleaned %>%
  mutate(across(where(is.character), as.factor))
cat("\nSummary After Cleaning:\n")
print(summary(final_data_cleaned))
summary(final_data_cleaned)

```
# Correlation Analysis, Outcome Variable Derivation, and Chi-Square Tests for Categorical Variables

```{r, echo=FALSE, warning=FALSE, message=FALSE}
cor_matrix <- cor(final_data_cleaned %>% select(where(is.numeric)), use = "complete.obs")
print(cor_matrix)

cat("Column Names in Dataset:\n")
print(colnames(final_data_cleaned))

if (!"outcome_variable" %in% colnames(final_data_cleaned)) {
  cat("Outcome variable does not exist. Deriving it...\n")
  
  final_data_cleaned$outcome_variable <- ifelse(final_data_cleaned$K2Q35A_1_YEARS > 5, "High", "Low")
}

final_data_cleaned$outcome_variable <- as.factor(final_data_cleaned$outcome_variable)

cat("Structure of the dataset:\n")
str(final_data_cleaned)

cat_vars <- names(final_data_cleaned %>% select(where(is.factor)))
cat("Running Chi-Square Tests on Categorical Variables:\n")
for (var in cat_vars) {
  if (var != "outcome_variable") {  
    cat("\nVariable:", var, "\n")
    print(chisq.test(table(final_data_cleaned[[var]], final_data_cleaned$outcome_variable)))
  }
}
```
# Correlation Heatmap and Visualization of Variable Distributions by Outcome

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Libraries
library(ggplot2)
library(reshape2)
library(dplyr)
library(caret)

cor_matrix <- cor(final_data_cleaned %>% select(where(is.numeric)), use = "complete.obs")
print(cor_matrix)

# Heatmap of Correlation Matrix
cor_melted <- melt(cor_matrix)
ggplot(cor_melted, aes(Var1, Var2, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0, limit = c(-1, 1)) +
  labs(title = "Correlation Matrix Heatmap", x = "Variables", y = "Variables", fill = "Correlation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

cat("Column Names in Dataset:\n")
print(colnames(final_data_cleaned))

if (!"outcome_variable" %in% colnames(final_data_cleaned)) {
  cat("Outcome variable does not exist. Deriving it...\n")
  final_data_cleaned$outcome_variable <- ifelse(final_data_cleaned$K2Q35A_1_YEARS > 5, "High", "Low")
}

final_data_cleaned$outcome_variable <- as.factor(final_data_cleaned$outcome_variable)
cat("Structure of the dataset:\n")
str(final_data_cleaned)

for (var in cat_vars) {
  if (var != "outcome_variable") {
    ggplot(final_data_cleaned, aes(x = .data[[var]], fill = outcome_variable)) +
      geom_bar(position = "fill") +
      labs(title = paste("Distribution of", var, "by Outcome"), x = var, y = "Proportion") +
      theme_minimal() +
      theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
      scale_fill_manual(values = c("High" = "red", "Low" = "blue")) -> plot
    print(plot)
  }
}

numeric_vars <- names(final_data_cleaned %>% select(where(is.numeric)))

for (var in numeric_vars) {
  ggplot(final_data_cleaned, aes(x = outcome_variable, y = .data[[var]], fill = outcome_variable)) +
    geom_boxplot() +
    labs(title = paste("Box Plot of", var, "by Outcome"), x = "Outcome", y = var) +
    theme_minimal() +
    scale_fill_manual(values = c("High" = "red", "Low" = "blue")) -> plot
  print(plot)
}
```

# Age Distribution Analysis by Outcome Category

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(ggplot2)

high_data <- final_data_cleaned %>% filter(outcome_variable == "High")

cat("Plotting Age Distribution for High Outcome...\n")
ggplot(high_data, aes(x = SC_AGE_YEARS)) +
  geom_histogram(binwidth = 1, fill = "blue", color = "black") +
  labs(title = "Age Distribution for High Outcome", x = "Age", y = "Count") +
  theme_minimal() +
  scale_y_continuous(breaks = seq(0, 5000, by = 1000))

low_data <- final_data_cleaned %>% filter(outcome_variable == "Low")

# Plot for "Low" outcome
cat("Plotting Age Distribution for Low Outcome...\n")
ggplot(low_data, aes(x = SC_AGE_YEARS)) +
  geom_histogram(binwidth = 1, fill = "red", color = "black") +
  labs(title = "Age Distribution for Low Outcome", x = "Age", y = "Count") +
  theme_minimal() +
  scale_y_continuous(breaks = seq(0, 100, by = 10))

```

# Random Forest Model Training, Evaluation, and Feature Importance Analysis with Visualization

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Libraries
library(caret)
library(randomForest)
library(pROC)
library(ggplot2)
library(reshape2)
set.seed(123)

train_index <- createDataPartition(final_data_cleaned$outcome_variable, p = 0.8, list = FALSE)
train_data <- final_data_cleaned[train_index, ]
test_data <- final_data_cleaned[-train_index, ]

# Train Random Forest Model
rf_model <- randomForest(outcome_variable ~ ., data = train_data, importance = TRUE, ntree = 500)

# 1. Variable Importance Plot
cat("Variable Importance Plot...\n")
varImpPlot(rf_model, main = "Random Forest Variable Importance")

# 2. Predictions and Confusion Matrix
rf_predictions <- predict(rf_model, newdata = test_data)
conf_matrix <- confusionMatrix(rf_predictions, test_data$outcome_variable)

# 3. Visual Confusion Matrix
conf_matrix_table <- as.data.frame(conf_matrix$table)
ggplot(data = conf_matrix_table, aes(x = Reference, y = Prediction, fill = Freq)) +
  geom_tile() +
  geom_text(aes(label = Freq), color = "white") +
  scale_fill_gradient(low = "blue", high = "red") +
  labs(title = "Confusion Matrix Heatmap", x = "Actual", y = "Predicted") +
  theme_minimal()

# 4. ROC Curve
if (is.factor(test_data$outcome_variable)) {
  rf_probabilities <- predict(rf_model, newdata = test_data, type = "prob")[, 2]
  roc_curve <- roc(as.numeric(test_data$outcome_variable) - 1, rf_probabilities) # Adjust for factor levels
  cat("ROC Curve...\n")
  plot(roc_curve, col = "darkgreen", lwd = 2, main = "ROC Curve for Random Forest")
  abline(a = 0, b = 1, lty = 2, col = "gray")
  cat("AUC:", auc(roc_curve), "\n")
}

# 5. Advanced Feature Importance Visualization
importance_df <- as.data.frame(importance(rf_model))
importance_df$Variable <- rownames(importance_df)
importance_df <- importance_df[order(importance_df$MeanDecreaseAccuracy, decreasing = TRUE), ]

ggplot(importance_df, aes(x = reorder(Variable, MeanDecreaseAccuracy), y = MeanDecreaseAccuracy)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Feature Importance by Mean Decrease in Accuracy", x = "Features", y = "Importance") +
  theme_minimal()

```


